---
title: "DATASCI 306: Lecture 4"
subtitle: "Graphics: Facets, Coordinate Systems"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
```

## Todays topic
 * More on ggplot layers
 * Coordinate systems
 * Facets

## Review

Quantitative vs. Categorical measurements in R

```{r}
class(c(1.0, 2.2, 3.14159)) # quantitative
class(mpg$class) # categorical 
class(mpg$cyl) # quantitative
class(factor(mpg$cyl)) # categorical makes more sense!
```

```{r}
aatemp <- read_csv("data/ann_arbor_weather.csv.gz")
```


## Geometries that expect categorical mappings

```{r}
ggplot(data = mpg, aes(x = cyl, y = cty)) +
  geom_boxplot()
```

## What we probably wanted

```{r}
ggplot(data = mpg, aes(x = factor(cyl), y = cty)) +
  geom_boxplot()
```



## Another overplotting example

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point()
```

Using jitter


```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_jitter()
```
Jitter isn't helping. How about adding `alpha` param to aes instead?

```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1/100)
```


## Facets: "sub grouping"

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_jitter() +
  facet_wrap(~ class)
```


Faceting sub groups the dataset based on each distinct value of the column `class` and then plots the 'x' and 'y' specified.

In R formulas, the `~` sign often indicates a relationship or a "by" clause. In the context of facet_wrap(), `~ class` means "facet (group the records) by the `class` variable."

## Forcing on one row

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_jitter() +
  facet_wrap(~ class, nrow = 1)
```

## Forcing two columns

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_jitter() +
  facet_wrap(~ class, ncol = 2)
```

## Facets can include everything plots have 

```{r warning = FALSE}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_jitter() +
  geom_smooth() + 
  facet_wrap(~ class) 
```


## Multiple factors

```{r warning = FALSE}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_jitter() +
  stat_smooth() + 
  facet_wrap(~ class + drv) 
```


## Two way facets (`facet_grid`)

```{r warning = FALSE}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_jitter() +
  stat_smooth() + 
  facet_grid(vars(drv), vars(cyl)) +
  labs(x = 'Cylinders', y = 'Drive train type')
```

Use vars() to supply variables from the dataset


## Exercise

Create a facet plot to visualize the relationship between price and carat, grouped by cut

```{r facet, exercise = TRUE}
ggplot(data = diamonds, aes(x = price, y = carat)) 
```

```{r facet-solution}
ggplot(data = diamonds, aes(x = price, y = carat)) +
  geom_point() +
  facet_wrap(~ cut)
```


## Coordinate Systems

Thus far we have been using the default **Cartesian coordinate system**.

There are many other ways in which we can connect our **position visual variables** to their actual locations on a plane.

These are called **coordinate systems** or **projections**.

## Coordinate systems for presentation

Why use a different coordinate system? One reason to provide a convenient or interesting **presentation**.

Base plot:
```{r}
overlap_boxplot <- ggplot(data = storms, aes(x = status, fill = status)) + geom_bar() 
overlap_boxplot
```

Drawback with this - labels overlap 

## Flipping coordinates

Solution for label overlap

```{r}
overlap_boxplot + coord_flip()
```

## Rotating labels
Of course, we could rotate the labels if we don't want to flip!

```{r}
overlap_boxplot + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```
axis.text.x: This specifies that you want to modify the properties of the x-axis tick labels.

element_text(): This function is used to define text properties.
angle: This argument sets the rotation angle in degrees. Common values include 45, 90 (for vertical labels), or other angles as needed.

hjust: This controls the horizontal justification of the text relative to the tick mark. A value of 1 aligns the right edge of the text with the tick mark (useful for rotated labels), while 0 aligns the left edge, and 0.5 centers the text.

vjust: This controls the vertical justification of the text. A value of 1 aligns the top edge of the text with the tick mark, 0 aligns the bottom edge, and 0.5 centers it vertically.

## Rearrange the bars

We can arrange the bars based on their count
```{r}
ggplot(data = storms) + geom_bar(aes(y = fct_infreq(status), fill = status)) + labs(fill = 'Status', y = 'Storm Status')
```

## Polar coordinates

Polar coordinates: express location as **angle** and **magnitude** from the origin.

```{r}
overlap_boxplot + coord_polar()
```

**`count` function**

Let us verify the counts 

```{r}
storms |> count(status)
```

## Temperature data for Ann Arbor, MI

```{r}
aatemp <- read_csv("data/ann_arbor_weather.csv.gz")
```

## Plotting temperature vs. time

```{r}
ggplot(data = aatemp, aes(x = DATE, y = TMAX)) +
  geom_line()
```

## Temperature Distribution and Variability

What exactly is the **seasonal effect** of temperature? 
```{r}
# remember: tables are LISTS of VECTORS:
aatemp$month <- month(aatemp$DATE)
ggplot(data = aatemp, aes(x = factor(month), y = TMAX)) +
  geom_violin(aes(fill = factor(month))) 
```

* High Variability: The greatest temperature swings (variability) occur during the winter and transitional spring/fall months.
* Predictability: Summer temperatures are more predictably high, while winter, spring and fall temperatures are much less predictable.


## Wrapping around using polar coordinates

```{r polar}
ggplot(data = aatemp, aes(x = factor(month), y = TMAX)) +
  geom_violin(aes(fill = factor(month))) +
  coord_polar()
```


## Temp difference from one day to next

```{r}
v = c(2, 4, 6, 7)
print(v)
diff(v)
```

```{r}
temp_diff <- diff(aatemp$TMAX)
length(temp_diff)
dim(aatemp)
aatemp$temp_diff <- abs(c(NA, temp_diff))
```

## Distributon of differences
```{r}
ggplot(data = aatemp, aes(x = temp_diff)) + geom_bar()
```

## Plotting when large jumps occur during the year

Let's define a large jump in maximum temperature as being when one day is more than 10 degrees different than the subsequent day.

```{r}
ggplot(data = filter(aatemp, temp_diff > 10), aes(x = yday(DATE))) +
  geom_histogram() + coord_polar()
```

## Other misceallaneous `ggplot` items

* Several other coordinate systems possible, in particular for representing geographic data.
* We saw using the `position` argument to jitter data. Other geometries have more options with jitter. See R documentation.
* We'll revisit more with labels, themes, and other presentation later in future classes.

